---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# centr

<!-- badges: start -->
[![codecov](https://codecov.io/github/ryanzomorrodi/centr/graph/badge.svg?token=TTGPNPXV8Z)](https://codecov.io/github/ryanzomorrodi/centr)
<!-- badges: end -->

A Package For Weighted and Unweighted Spatial Centers

## Installation

You can install the development version of centr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ryanzomorrodi/centr")
```

## Usage

The main functions are `mean_center` and `median_center`. They were designed for calculation of population weighted centroids, but can be extended to other forms of analyses.

Mean center calculates the geographic average center. One can specify the groups to calculate individual centers for groups and weights for each individual point. It is analagous to the [ArcGIS Pro Mean Center](https://pro.arcgis.com/en/pro-app/latest/tool-reference/spatial-statistics/mean-center.htm) tool. 

```{r mean_example}
library(centr)
df <- data.frame(
  lon = c(20, 50, 30, 80, 10),
  lat = c(25, 70, 30, 50, 30),
  grp = c("a", "b", "a", "b", "a"),
  wt = c(1, 5, 1, 3, 2)
)
x <- sf::st_as_sf(df, coords = c("lon", "lat"), crs = 4326)

mean_center(x, group = "grp", weight = "wt")
```

Median center iteratively calculates the point that minimizes distance to all features. One can specify the groups to calculate individual centers for and weights for each individual point. It is analagous to the [ArcGIS Pro Median Center](https://pro.arcgis.com/en/pro-app/latest/tool-reference/spatial-statistics/median-center.htm) tool.

```{r median_example}
library(centr)
df <- data.frame(
  lon = c(20, 50, 30, 80, 10),
  lat = c(25, 70, 30, 50, 30),
  grp = c("a", "b", "a", "b", "a"),
  wt = c(1, 5, 1, 3, 2)
)
x <- sf::st_as_sf(df, coords = c("lon", "lat"), crs = 4326)
x_transformed <- sf::st_transform(x, crs = "ESRI:102003")

median_center(x_transformed, group = "grp", weight = "wt")
```
