---
title: "Typical Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typical Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package was designed to be used to easily calculate population weighted centroids within R. While there are many scenarios in which you may require a weighted centroid, I present the following as an typical workflow for my usage.

For this example we are going to be calculating the population weighted centroids for North Carolina counties. As such, we require population and geographic data at a more granular geographic classification. We will use tracts. 

## Downloading Data
We can download census population estimates and geographic layer files using the ever convenient `tidycensus` package. `sf` also has a North Carolina counties file, which we will use later, but you can always download census geometries directly from the census using `tigris`.
```{r setup, results = 'hide', message = FALSE, warning = FALSE}
library(centr)
library(sf)
library(tidycensus)

NC_tracts <- get_decennial("tract", state = "NC", "P1_001N", year = 2020, geometry = TRUE)
NC_counties <- st_read(system.file("shape/nc.shp", package = "sf")) 
```

## Setting Up
We will need fields that tells us what county each tract is in and how many people live within each tract. Thankfully, the first 5 digits of the GEOID, uniquely identify the county a tract is in, which means that we can easily create a new field with this ID. The value field also gives us our weights.
```{r}
NC_tracts <- NC_tracts |>
  transform(GEOID_county = substring(GEOID, 1, 5)) |>
  subset(select = c(GEOID_county, value))
```

Let's use `mean_center()` to get our county geometries!
```{r, error = TRUE}
mean_center(NC_tracts, group = "GEOID_county", weight = "value")
```

Oh, it looks like we forgot one thing. At least one of our tracts has an empty geometry. Before running `mean_center()`, we must filter out empty geometries. 
```{r}
NC_tracts <- subset(NC_tracts, !st_is_empty(NC_tracts))
```

## Calculation Mean Centers
Now, we can calculate our mean centers!
```{r}
NC_county_means <- mean_center(NC_tracts, group = "GEOID_county", weight = "value")
NC_county_means
```

Let's see how it looks.
```{r}
plot(st_geometry(NC_counties))
plot(st_geometry(NC_county_means), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

## Calculating Median Centers
We can also calculate median centers, which is the point that minimizes distance to all geometries. `median_center()` only supports projected coordinates, so let's project to the North Carolina state plane.
```{r}
NC_tracts_proj <- st_transform(NC_tracts, crs = "EPSG:32119")

NC_county_medians <- NC_tracts_proj |>
  median_center(group = "GEOID_county", weight = "value")
NC_county_medians
```

Let's see how it looks.
```{r}
NC_counties_proj <- st_transform(NC_counties, crs = "EPSG:32119")

plot(st_geometry(NC_counties_proj))
plot(st_geometry(NC_county_medians), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

And that's the whole game, so go out and try it for yourself!